Analysing Melbourne CBD pedestrian counts data with a focus on imputation methods

### 2017-10-21 thesis_draft/00-abstract.Rmd
bookdown update - need to fix formatting, improve introduction, write abstract, finish off lit review and conclusion

---
title: "Analysing Melbourne CBD pedestrian counts data with a focus on imputation methods"
author: "Gavin Chin"
date: "21 September 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE,
                      message = FALSE, warning = FALSE,
                      fig.pos = 'h')
library(tidyverse)
library(lubridate)
library(foreach)
library(doSNOW)
library(rwalkr)
library(pander)
library(RColorBrewer)
library(gridExtra)
library(viridis)
library(ggmap)
library(broom)
set.seed(12345)
```

```{r data_loading, warning=F, message=F}
ped_data <- NULL
if(file.exists("../data/ped_df.csv")){
  ped_data <- read_csv("../data/ped_df.csv")
  ped_data$X1 <- NULL
  }

 ## ped_data <- read_csv("http://209.148.91.227/files/ped_df_28sep17.csv")

## update data
last_df_date <- ifelse(file.exists("../data/ped_df.csv"),
                        ped_data$Date %>% sort() %>%
                        tail(1),
                        as.Date("2013-12-31")) + 1 %>% as.Date(origin = "1970-01-01")

if(last_df_date != today(tzone = "Australia/Melbourne")){
  new_data <- rwalkr::walk_melb(from = last_df_date, to = today()-1)
  ped_data <- rbind(ped_data, new_data)
  write_csv(ped_data, "../data/ped_df.csv")
  }

ped_data$Sensor_ID <- ped_data$Sensor
ped_data$Hourly_Counts <- ped_data$Count
ped_data$Date_Time <- ymd_h(paste(ped_data$Date, ped_data$Time, sep = " "))
ped_data$Day <- lubridate::wday(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Month <- lubridate::month(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Time <- as.factor(ped_data$Time)

## long to wide
ped_data %>% select(Sensor_ID, Count, Date_Time, Time, Month, Day) %>% 
  spread(Sensor_ID, Count) -> dfa

pub_hday14 <- read.csv("http://data.gov.au/dataset/b1bc6077-dadd-4f61-9f8c-002ab2cdff10/resource/56a5ee91-8e94-416e-81f7-3fe626958f7e/download/australianpublicholidays-201415.csv---australianpublicholidays.csv.csv")
pub_hday15 <- read.csv("http://data.gov.au/dataset/b1bc6077-dadd-4f61-9f8c-002ab2cdff10/resource/13ca6df3-f6c9-42a1-bb20-6e2c12fe9d94/download/australianpublicholidays-201516.csv")
pub_hday16 <- read.csv("http://data.gov.au/dataset/b1bc6077-dadd-4f61-9f8c-002ab2cdff10/resource/a24ecaf2-044a-4e66-989c-eacc81ded62f/download/australianpublicholidays-201617.csv")
pub_hdays <- rbind(pub_hday14, pub_hday15, pub_hday16)

pub_hdays$Date <- ymd(pub_hdays$Date)
pub_hdays$Month <- lubridate::month(pub_hdays$Date, label = TRUE, abbr = F)
pub_hdays$VIC <- 0
pub_hdays$VIC[grep(glob2rx("*VIC*"), pub_hdays$`Applicable.To`)] <- 1
pub_hdays$VIC[grep("NAT", pub_hdays$`Applicable.To`)] <- 1

dfa <- dfa %>% mutate(IsHDay = (date(Date_Time) %in% pub_hdays$Date[pub_hdays$VIC == 1]))

dfa$DayType <-       ifelse(dfa$Day == "Sunday", "Sunday",
                     ifelse(dfa$Day == "Saturday", "Saturday",
                    ifelse(dfa$Day == "Monday", "Monday",
                     ifelse(dfa$Day == "Friday", "Friday",
                                  "Midweek"))))
dfa$DayType[dfa$IsHDay == 1] <- "Holiday"
dfa$HDay <- ifelse(dfa$IsHDay == 1, "Holiday", dfa$Day)
original_df <- dfa
dfa <- filter(dfa, Date_Time < "2017-01-01")
ori_2 <- filter(original_df, Date_Time < "2017-01-01")
ori_ts <- filter(original_df, Date_Time > "2016-12-31")
ori_ts2 <- ori_ts
```
## Background

#### The Data: City of Melbourne Pedestrian Data

@@

The dataset being investigated can be obtained from the official City of Melbourne's open data[^2]. The pedestrian data is in the form of hourly pedestrian counts for `r length(unique(ped_data$Sensor_ID))` sensor locations. In total, there is `r nrow(ped_data)` observations in the dataset, representing `r nrow(dfa)` hours of data. It is available in `.csv` format, allowing for easy data import into R. We can also access the pedestrian data using the the `rwalkr` package by Earo Wang[^3]. This package allows R to import the data from `data.melbourne.vic.gov.au` which is updated monthly, or from the data source which is used by `pedestrian.melbourne.vic.gov.au` which is updated daily.

```{r}
ped_data %>% select(Sensor, Hourly_Counts) %>% split(.$Sensor)  %>%
            map(is.na) %>% map(sum) %>% unlist -> sensor_missvals
```

Not even an single sensor location has complete data for the period between between 1/1/2014 and `r paste(day(today()), month(today()), year(today()), sep = "/")`.

### Literature Review
<!-- #### IMPUTATION OF MISSING CLASSIFIED TRAFFIC DATA DURING WINTER SEASON - _Hyuk-Jae Roh, Satish Sharma, Prasanta K. Sahu_   -->

<!-- http://ir.lib.uwo.ca/cgi/viewcontent.cgi?article=1250&context=csce2016   -->
<!-- - traffic counters with 40% to 60% missing values -->
<!-- - heuristic methods -->
<!--   - using "good historical values" or historical average values resulted -->
<!--     in MARE (mean absolute relative error) reaching up to 80% -->
<!--   - using moving average values worked better than other heuristic methods -->
<!--   - all heuristic methods inhenrently cannot reflect sudden fluctuations during abnormal periods -->
<!-- - pattern matching methods -->
<!--   - find "candidate" volume pattern to compare to "study curve" and use the candidate which best matches the study curve -->
<!-- - ARIMA -->
<!--   - did not work well with multiple seasonalities -->
<!-- - AI methods -->
<!--   - genetic algorithms (GAs) and artificial neural networks (ANNs) -->

<!-- Method used in the study was kNN with an interaction using a weather model, which they found to work best with large variations in traffic during winter -->
<!-- (severe weather conditions affecting traffic volume)   -->

Reviewing current research on imputation of missing values in traffic data, a paper on imputation of missing classified traffic data during winter season (Imputation of Missing Classified Traffic Data During Winter Season - _Hyuk-Jae Roh, Satish Sharma, Prasanta K. Sahu_) stated some methods which were found to be poor for imputation. The data they were attempting imputation for were traffic counters located on the highway network in Alberta, Canada with between 40% and 60% missing data.

Replacement with "good" historical values, or using historical average values was found to have resulted in very poor fit with high MARE/MARD (mean absolute relative error/difference). They did find that using a moving average worked the best out of all the heuristic methods used. The largest problem with heuristic methods, however, is the inherent inability to reflect sudden fluctuations/shocks during abnormal periods. With such a large proportion of missing values, pattern matching methods were investigated. This involves comparing the _study curve_, the pattern at the counter which is to be imputed, to candidate patterns (patterns at other locations).

For the purposes of that paper, it was found to be inappropriate with the data being analysed.  However, unlike _Roh et al._ where patterns were compared to traffic volumes in different jurisdictions (large geographical distance), the geographic distances between the sensor locations in the Melbourne CBD pedestrian data are much smaller. Another difference is the randomness of the missing values, where the periods of missing data run longer in the Melbourne CBD pedestrian data compared to that in the Alberta traffic data. They proposed using non-parametric estimation methods with k-Nearest Neighbours for their imputation, although this is unlikely to work well with the lack of data at certain sensor locations in the Melbourne CBD Pedestrian data.

<!-- #### Flexible and Robust Method for Missing Loop Detector Data Imputation - _K. C. Henrickson, Y. Zou, and Y. Wang_   -->

<!-- http://docs.trb.org/prp/15-5804.pdf   -->
<!-- - also use kNN -->

<!-- #### Key Factors of K-nearest Neighbors Nonparametric Regression in Short-time Traffic Flow Forecasting -->
<!-- http://www.springer.com/cda/content/document/cda_downloaddocument/9789462391017-c2.pdf?SGWID=0-0-45-1492476-p177147492 -->

<!-- #### K-NEAREST NEIGHBOR ALGORITHM FOR UNIVARIATE TIME SERIES PREDICTION -->

<!-- http://webbut.unitbv.ro/BU2012/Series%20III/BULETIN%20III/Sasu%20A.pdf -->

## Methodology: Imputation of Missing Values

### Basic GLM Approach

The first method used to impute missing values is to fit generalised linear models (GLMs) at each sensor location. Estimating a GLM with a Quasi-Poisson error distribution, where specify the model as:
$$E \left[\text{HourlyCounts} \vert \text{Sensor}, \text{HDay}, \text{Time} \right] = \exp (\mu_{\text{Sensor}, \text{HDay}, \text{Time}}) $$
$$\text{where } \quad \mu_{\text{Sensor}, \text{HDay}, \text{Time}} \sim \text{Time}\times \text{HDay} \tag{1}$$
`Time` is the time of the day, while `HDay` is the day of the week, with an additional factor level for public holidays. 
Both these variables are categorical/factors. We cannot treat time of the day as an integer or numeric because it does not have a linear relationship with counts.s

We use the Quasi-Poisson error distribution as opposed to normal/Gaussian errors due to the response variables being count data. The Poisson distribution allows for only non-negative integer values, while estimating a Quasi-Poisson model estimates an additional parameter for overdispersion. This allows more flexibility in the model by allowing the assumption that $E[Y]=Var[Y]=\mu$ to be relaxed. Instead, the Quasi-Poisson distribution allows for $Var[Y] = \theta \mu$.

```{r}
## Base GLM models

### code from:
## http://www.win-vector.com/blog/2014/05/trimming-the-fat-from-glm-models-in-r/
stripGlmLR = function(cm) {
  cm$y = c()
  cm$model = c()

  cm$residuals = c()
  cm$fitted.values = c()
  cm$effects = c()
  cm$qr$qr = c()  
  cm$linear.predictors = c()
  cm$weights = c()
  cm$prior.weights = c()
  cm$data = c()

  cm$family$variance = c()
  cm$family$dev.resids = c()
  cm$family$aic = c()
  cm$family$validmu = c()
  cm$family$simulate = c()
  attr(cm$terms,".Environment") = c()
  attr(cm$formula,".Environment") = c()

  cm
}
#######################################
```

```{r}
cl <- makeCluster(2)
registerDoSNOW(cl)
fitted_base <- list()
models_base <- foreach(i = unique(ped_data$Sensor_ID)) %dopar% {
    tr_dat <- data.frame(dfa[, i],  HDay = dfa$HDay, Time = dfa$Time)
    colnames(tr_dat)[1] <- "Hourly_Counts"
    model <- glm(Hourly_Counts ~ HDay*Time, data = tr_dat, family = quasipoisson())
    return(stripGlmLR(model))
  }
names(models_base) <- unique(ped_data$Sensor_ID)
# ped_models[[31]]$xlevels <- ped_models[[1]]$xlevels
for (i in unique(ped_data$Sensor)){  
  tr_dat <- data.frame(dfa [, i],  HDay = dfa$HDay, Time = dfa$Time)
  options(na.action = na.pass)
  fitted_base[[i]] <- predict.glm(models_base[[i]], newdata = tr_dat, type = "response")
  options(na.action = na.omit)
  }

```

While this works well with a small proportion of missing values, it also requires a large number of observations. Specifically, with this paramaterisation, we have $23 + 7 + (23 \times 7) + 1=192$ coefficients to estimate. Another disadvantage of using this model is the lack of robustness to outliers due to the sensitivity of some parameters. This becomes particularly problematic at sensor locations which have been recently installed, where there is a lack of historical data.

### Improved imputation algorithm using small-large split approach

The proposed alternative approach to imputing the missing values in the data focuses on improving the models used at locations with a large proportion of missing values. A potential threshold value which could be used to class a sensor as having a "large" proportion of missing values vs a "small" proportion of missing values is 10%. At locations with a "small" proportion of missings, we can continue to use the GLM quasi-possion regression specified in $(1)$. At the locations with a large proportion of missing values, we need to use information from neighbouring sensors.

The first issue which needs to be addressed is the potential of values which are actually missing due to sensor failure or malfunction having a zero count. This issue will cause bias in the estimates, as well as affect the classification of a sensor. Of course, we cannot simply classify all zero counts as `NA`, as true zero values are possible.

```{r}
n_obs <- nrow(dfa)
threshold <- 0.1
threshold_miss_hrs <- 6

run_lengths <- list()

for(i in unique(ped_data$Sensor)){
    dfa[is.na(dfa[, i]), i] <- 0
    na_length_check <- rle(as.numeric(unlist(dfa[,i])))
    maybeNA <- rep((na_length_check$lengths > threshold_miss_hrs), times = na_length_check$lengths)
    rl <- dfa[, i] %>% unlist %>% rle()
    dfa[maybeNA, i] <- NA
    defsNA <- is.na(ori_2[, i])
    dfa[defsNA, i] <- NA
    run_lengths[[i]] <- rl
  }

ori_2 <- dfa

run_lengths %>% lapply(., `[[`, 1) %>%
    unlist -> sequences

pander(summary(sequences[sequences > 1]))
```
Table: Summary statistics of the sequence lengths of repeated values

Instead, a simple check we can implement is to look for long sequences of zero counts. For example, we may want to classify any sequence of `Hourly_Counts = 0` running for longer than 24 hours as being considered `NA`. From the summary of lengths of repeated values (which are repeated at least once), we see that the distribution of the lengths are very positively skewed with a mean length of `r round(mean(sequences[sequences > 1]), 2)` hours of missing data. Classifying any sequence of repeated values, typically 0, which has a length greater than 6 as `NA` should allow true zero counts to avoid being misclassified as `NA`. We select the length of 6 hours as it would be unreasonable to assume any sensor location to have the exact same number of pedestrian counts over such a long period.

After we have replaced questionable zero values with the value `NA`, we can calculate the proportion of missing values for each sensor location. From these calculated proportion, we can classify sensors as either having a "large" or "small" proportion of missing values, where large is defined as $\mathtt{NAprop}_{sensor} > \mathtt{threshold}$ and `threshold = 0.1`.

```{r na_prop_compute}
na_prop <- numeric()

for (i in unique(ped_data$Sensor)){
  na_prop[i] <- sum(is.na(dfa[, i])) / n_obs
}

na_prop_pre <- numeric()

for (i in unique(ped_data$Sensor)){
  na_prop_pre[i] <- sum(is.na(ori_2[, i])) / n_obs
}

large_miss_sensors <- names(na_prop[na_prop > threshold])
small_miss_sensors <- setdiff(names(na_prop), large_miss_sensors)
```

```{r naprop_hist, fig.width=9, fig.height=4, fig.cap="Histograms of the proportions of missing values at each sensor during 2014 - 2016 when calculated before and after correction for false zero counts. The distributions of proportion of missing values at each sensor is significantly different after correcting for false zero counts in the data. This emphasises the necessity to correct for false zero counts as it will affect "}
theme_g <- theme_minimal() +
           theme(plot.title = element_text(hjust = 0.5,
                                           size = 10))
p1 <- ggplot() + geom_histogram(aes(x = na_prop*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
p2 <- ggplot() + geom_histogram(aes(x = na_prop_pre*100), bins = 12) +
          xlab("Percentage of Missing Values (%)") +
          ylab("Number of Sensors") + theme_g +
          geom_vline(xintercept = 10, size = 2,
                     colour = 'red') +
          geom_label(aes(label = "10% Threshold", x = 16, y = 25),
                    colour = "white", fill = 'black', alpha = 0.4) +
          scale_y_continuous(limits = c(0,35))
grid.arrange(p2, p1, nrow = 1)
```

Because the models for the sensors with a large proportion of missing values rely on having neighbouring sensors having complete cases (no missing values), we want to impute these values first. This will then allow us to have as much information to be available for the models which use neighbouring sensors to to train on.

For sensors with a small proportion of missing values, we use a GLM quasipoisson model. To improve on the previous model used, we use the model:
$$E \left[\text{HourlyCounts} \vert  \text{Sensor}, \text{Month}, \text{DayType}, \text{Time} \right] = \exp (\mu_{\text{Sensor}, \text{Month}, \text{DayType}, \text{Time}}) $$

$$\text{where } \quad \mu_{\text{Sensor}, \text{Month}, \text{DayType}, \text{Time}} \sim \text{Month} + \text{Time}\times \text{DayType} \tag{2}$$

```{r alg_glm}
models <- list()
fitted_data <- list()
dfa2 <- dfa
p <- progress_estimated(length(unique(ped_data$Sensor)))
for(i in small_miss_sensors){
  p$tick()$print()
  options(na.action = na.omit)
  # if(max(na_length_check$lengths) > max_miss_hrs) 
  # {
  #   model <- NULL
  # }

  sensor_df <- unlist(dfa[, i])

  model <- glm(sensor_df ~  Month + DayType*Time, family = 'quasipoisson', data = dfa)
  models[[i]] <- (stripGlmLR(model))
}
```

```{r alg_glm_impute}
  options(na.action = na.pass)
for(i in small_miss_sensors){
  ts_dat <- data.frame(Month = dfa$Month, DayType = dfa$DayType, Time = dfa$Time)
  fitted_data[[i]] <- predict.glm(models[[i]],newdata = ts_dat, type = "response")
}
  options(na.action = na.omit)
  for(i in small_miss_sensors){
    dfa2[is.na(dfa[, i]), i] <- fitted_data[[i]][is.na(dfa[, i])]
    }

```

Instead of `HDay`, we use `DayType` which classifies Tuesday, Wednesday and Thursday as a single factor level, Midweek. We do this as the daily patterns on these weekdays are similar. This reduces the number of parameters to estimate by 46. A result of this is we also have more robust estimates for the midweek days, being less sensitive to outliers. Because this model will be estimated for sensor locations with few missing values,  we have enough data to add in month as an additive effect. This will help capture the annual seasonality.

Using these estimated models, we impute the missing values to produce complete data at all these sensor locations. For the locations with high proportions of missing values can be predicted using the now complete data from all the locations which had a small proportion of missing values. Using a threshold of 10%, this gives us `r length(small_miss_sensors)` sensor locations to use.

```{r, fig.width=3, fig.height=3, fig.align='center', fig.cap = "Map showing the algorithm selecting Flinders St Station Underpass and Sandridge Bridge as neighbours to be used to predict the pedestrian counts at Southbank"}
ped_loc <- read_csv("../data/Pedestrian_sensor_locations.csv")
lats <- ped_loc$Latitude
lons <- ped_loc$Longitude
ids <- ped_loc$`Sensor Description`

{
ids[ids == "Flinders Street Station Underpass"] <- "Flinders St Station Underpass"
ids[ids == "Flinders St-Spark La"] <- "Flinders St-Spark Lane"
ids[ids == "Queen St (West)"] <-  "Queen Street (West)"
ids[ids == "The Arts Centre"] <- "Vic Arts Centre"
ids[ids == "Lonsdale St-Spring St (West)"] <- "Spring St-Lonsdale St (South)"
ids[ids == "Lygon St (East)"] <- "Lygon Street (East)"
ids[ids == "QV Market-Elizabeth St (West)"] <- "QV Market-Elizabeth (West)"
ids[ids == "Melbourne Convention Exhibition Centre"] <- "Convention/Exhibition Centre"
ids[ids == "St Kilda Rd-Alexandra Gardens"] <- "St. Kilda-Alexandra Gardens"
}
loc <- data.frame(lons, lats, ids)
rownames(loc) <- ids
loc <- loc %>% mutate(cl = ifelse((loc$ids %in% small_miss_sensors),
                                  "Small", "Large"))
rownames(loc) <- loc$ids

melb3 <- get_map(location = c(loc["Southbank",]$lons,
                              loc["Southbank",]$lats),
                 zoom = 17, maptype = "roadmap")

sens_lines_lon <- as.numeric(c(loc["Flinders St Station Underpass",]$lons,
                loc["Southbank",]$lons,
                loc["Sandridge Bridge",]$lons))

sens_lines_lat <- as.numeric(c(loc["Flinders St Station Underpass",]$lats,
                loc["Southbank",]$lats,
                loc["Sandridge Bridge",]$lats))

sens_lines <- data.frame(lon = sens_lines_lon,
                         lat = sens_lines_lat)
options(na.action = 'na.omit')
ggmap(melb3) + geom_line(aes(x = lon, y = lat),
                        data = sens_lines, size = 2,
                        colour = viridis(3)[2], alpha = 0.5) +
               geom_point(aes(x = lons, y = lats, colour = cl),
                          size = 6,
                          data = loc) +
               scale_colour_viridis(discrete = TRUE,
                                    name = "Sensor NA Prop.",
                                    begin = 0.5, end = 0) +
               geom_point(aes(x = loc["Southbank",]$lons,
                              y = loc["Southbank",]$lats),
                          size = 12, colour = viridis(3)[2]) +
               geom_label(aes(label = ids, x = lons, y = lats,
                              fill = cl),
                         fontface = 'bold', colour = 'white',
                         data = loc, size = 3,
                         nudge_y = -0.000325) +
              scale_fill_viridis(discrete = TRUE,
                                  name = "Sensor NA Prop.",
                                  begin = 0.5, end = 0) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank(),
                    legend.position = c(.8,0.2))
```

For each of the sensor locations with high proportions of missings, we need to decide which sensor to use for prediction. A simple method of finding a geographical closest neighbour is to take the haversine/great-circle distance between the sensor to be imputed and all the possible candidates.
```{r, message=F}
# Calculates the geodesic distance between two points specified by radian latitude/longitude using the
# Spherical Law of Cosines (slc)
# source: https://www.r-bloggers.com/great-circle-distance-calculations-in-r/
gcd.slc <- function(long1, lat1, long2, lat2) {
  R <- 6371 # Earth mean radius [km]
  d <- acos(sin(lat1)*sin(lat2) + cos(lat1)*cos(lat2) * cos(long2-long1)) * R
  return(d) # Distance in km
}

## name fix dictionary
{
ids[ids == "Flinders Street Station Underpass"] <- "Flinders St Station Underpass"
ids[ids == "Flinders St-Spark La"] <- "Flinders St-Spark Lane"
ids[ids == "Queen St (West)"] <-  "Queen Street (West)"
ids[ids == "The Arts Centre"] <- "Vic Arts Centre"
ids[ids == "Lonsdale St-Spring St (West)"] <- "Spring St-Lonsdale St (South)"
ids[ids == "Lygon St (East)"] <- "Lygon Street (East)"
ids[ids == "QV Market-Elizabeth St (West)"] <- "QV Market-Elizabeth (West)"
ids[ids == "Melbourne Convention Exhibition Centre"] <- "Convention/Exhibition Centre"
ids[ids == "St Kilda Rd-Alexandra Gardens"] <- "St. Kilda-Alexandra Gardens"
}
loc <- data.frame(lons, lats, ids)
rownames(loc) <- ids
```

```{r alg_lm}
for (i in large_miss_sensors){
  # p$tick()$print()
  sensor_dists <- numeric()
  for (j in small_miss_sensors){
      dists <- gcd.slc(loc[i,]$lons, loc[i,]$lats, loc[j,]$lons, loc[j,]$lats)
      sensor_dists[j] <- dists
  }
  names(sensor_dists) <- small_miss_sensors
  closest_sensor <- sensor_dists %>% sort %>%
                    head(2) %>% names
  close_df_scale <- dfa2[, closest_sensor[1]] %>% unlist %>% scale
  close_df_scale2 <- dfa2[, closest_sensor[2]] %>% unlist %>% scale

  sensor_df <- dfa2[, i] %>% unlist

  tr_dat <- data.frame(sensor_df, close_df_scale, close_df_scale2, Time = dfa$Time)
  options(na.action = na.omit)
  model <- glm(sensor_df ~ Time*close_df_scale + Time*close_df_scale2,
               data = tr_dat, family = quasipoisson())
  models[[i]] <- stripGlmLR(model)

  options(na.action = na.pass)
  fitted_data[[i]] <- predict(model, newdata = tr_dat, type = "response")
  options(na.action = na.omit)

  dfa2[is.na(dfa[, i]), i] <- fitted_data[[i]][is.na(dfa[, i])]

}

```

Using the two geographically closest sensors with complete cases, we use another GLM specified as:
$$E \left[ \text{HourlyCounts}_{\text{sensor}} \vert \text{Time}, \text{HourlyCounts}_{\text{neighbours}} \right] = \mu_{\text{Time}, \text{HourlyCounts}_{\text{neighbours}}}$$
$$\text{where } \quad  \mu_{\text{Time}, \text{HourlyCounts}_{\text{neighbours}}} \sim \text{Time} \times \text{HourlyCounts}_{\text{neighbour 1}}^{SC} + \text{Time} \times \text{HourlyCounts}_{\text{neighbour 2}}^{SC}$$

We would expect that the amount of people passing through the neighbouring sensors will be a strong predictor of the pedestrian counts as they are likely to also pass through. Using geographical neighbours, we can capture rare events effectively. Any large shocks to the counts at the neighbouring sensors will have an effect on the counts to be imputed.

```{r, echo = FALSE, message=FALSE, warning=F, include=FALSE}
i <- "Southbank"
  sensor_dists <- numeric()
  for (j in small_miss_sensors){
      dists <- gcd.slc(loc[i,]$lons, loc[i,]$lats, loc[j,]$lons, loc[j,]$lats)
      sensor_dists[j] <- dists
  }
  names(sensor_dists) <- small_miss_sensors
  closest_sensor <- sensor_dists %>% sort %>%
                    head(2) %>% names
  dfa3 <- original_df %>% filter(Date_Time > "2017-01-01")
  close_df <- dfa3[, closest_sensor[1]] %>% unlist
  close_df2 <- dfa3[, closest_sensor[2]] %>% unlist

  close_dfsc <- scale(close_df)
  close_dfsc2 <- scale(close_df2)

  sensor_df <- dfa3[, i] %>% unlist
  sensor_dfsc <- scale(sensor_df)

  tr_dat <- data.frame(sensor_df, close_df, close_df2,
                       Time = as.integer(dfa3$Time),
                       Date_Time = dfa3$Date_Time,
                       HDay = dfa3$HDay) %>%
            gather(key = Sensor, value = Count, sensor_df:close_df2)

  tr_dat_sc <- data.frame(sensor_dfsc, close_dfsc, close_dfsc2,
                       Time = as.integer(dfa3$Time),
                       Date_Time = dfa3$Date_Time,
                       HDay = dfa3$HDay) %>%
            gather(key = Sensor, value = Count, sensor_dfsc:close_dfsc2)

p1 <- ggplot(tr_dat %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + theme_minimal() + ggtitle("Unscaled Neighbours") +
      theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

p2 <- ggplot(tr_dat_sc %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + ggtitle("Scaled/Stadardised Neighbours") +
      ylab("Standard Deviations") + theme_minimal() +
      theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

```{r, echo = FALSE, message = FALSE, warning = F, fig.width = 8, fig.height = 4}
grid.arrange(p1, p2, ncol = 1)
```

In order to use a mixture of the patterns from two different sensors, we need to use the scaled counts at the neighbouring sensors to avoid the magnitude of the counts having an effect on the predicted counts. We scale the counts by standardising to $\mu = 0, \sigma^2 = 1$. Because the covariance with neighbouring sensors may vary over time, we also add an interaction term between the neighbouring sensor counts and the time of the day.

## Implementation of improved imputation algorithm

The exact code used for this paper can be found on http://github.com/gavinchin/honours2017

### False zero-counts correction

The false zero-counts correction is performed in R using the run length encoding function, `rle()`. This function computes the length and values of runs of equal values in a vector. Unfortunately, this function does not recognise `NA` in the its computation, so all NA values are initially recoded to take the value 0 at each sensor. This results in all missing values and true zero counts to take the value zero when running `rle()`. With the returned run lengths, all sequences of repeated values longer than the defined length of 6 hours are replaced with `NA`. Any values which were also originally missing, but were not missing for longer than 6 hours (and hence are still coded as 0), are recoded to be `NA`. It must be noted that sequences of repeated values not equal to zero which run for longer that 6 hours would be incorrectly replaced with `NA`. This is highly unlikely, however, as 6 hours of the exact same pedestrian counts would be extremely rare.

### Caculation of proportion of missing values and classification of sensors

In the working data, a wide format is used. This makes it easier to calculate missing value proportion at each sensor location, as it is the proportion of missing values in the column for each sensor. Using `is.na()` returns a logical vector indicating whether the sensor count is missing/`NA` or not. The sum of the instances where `is.na = TRUE` divided by the hours in the period of 2014-2016 is computed to get the proportion of missing values.

These proportions are saved for each sensor and then classified as either "large" or "small" based on a threshold value for the proportion. A list of sensors with large proportions of missing data and a list of sensors with small proportions of missing data is generated.

### Estimation of GLM at sensor locations with a small proportion of missing values

A for loop is used to estimate the GLM specified:
```{r, echo = TRUE, eval = FALSE}
glm(dfa[, i] ~ Month + DayType*Time, data = dfa,
    family = quasipoisson())
```
where `dfa` is the object name given to the data frame containing the training data after the false zero-count correction, and `i` is the $i^{th{}}$ sensor in the list of sensors with a small proportion of missing values. These models are stored in a list, forming a list of models. An alternative approach to using a for loop in R is splitting the data into separate data frames for each sensor location (forming a list of data frames) and applying the `glm()` function on each data frame in the list using the function `purrr::map()`. However, when this was attempted, performance issues were encountered due to the large size of the estimated model objects in R. This was not an issue when a for loop was used. Additional code for removing unnecessary elements of the GLM model object before storing to the list of models, reduucing the memory usage.

### Imputation at sensor locations with a small proportion of missing values

The predicted values are obtained using `predict.glm()`, using the option `type = "response"` to have predicted values returned predicted counts. The default option returns the predicted values of $\hat{\lambda}$ in the quasipoisson process, not the counts, $\exp (\hat{\lambda})$.  
Missing values are then replaced using the imputed values, generating a data frame of complete data as sensor locations for with a small proportion of missing values to be used for prediction at sensors with a large proportion of missing values.

### Estimation procedure at sensor locations with a large proportion of missing values
The following steps are performed on each sensor with a large proportion of missing values inside a for loop.

##### Finding neighbouring sensors
Using location data of the sensors provided by the City of Melbourne, a data frame containing the latitude and longitude coordinates of all the sensors which were classified as having a small proportion of missing values (and now have imputed, complete data), which are the candidates to be neighbours used for prediction.  
Using the Great Circle/Haversine distance equation, implemented in the function `gcd.slc()`, an approximation of distance between each of the candidate sensors is calculated in and stored as a vector. Sorting by ascending order (using `sort()`) and taking the two smallest distances with `head(., 2)` returns the two geographically closest candidate neighbours.

##### Extracting neighbouring sensors counts and standardising

Using the sensor location names, the neighbouring sensors counts are extracted from the imputed training data and each vector is standardised to $\mu = 0, \sigma^2 = 1$ using the base R function `scale()` then stored.

##### GLM estimation

The GLM for sensor locations with a large proportion of missing values also uses `glm()` as follows:
```{r, echo = TRUE, eval = FALSE}
glm(dfa[, i] ~ Time*close_dfsc + Time*close_dfsc2,
    data = dfa, family = quasipoisson())
```
where `dfa[, i]` is the vector of pedestrian counts for the sensor being estimated, `Time` is the factor level for the hour of the day and `close_dfsc` and `close_dfsc2` are the standardised pedestrian counts at two geographically closest sensors.  
Like the GLM for sensor locations with a small proportion of missing values, the model object is slim down to remove unnecessary elements to reduce memory usage. It is also noteworthy that the computational burden of the procedure for imputation using neighbouring sensors is significantly less due to less parameters being estimated.

##### Replacing missings with imputed values  
Unlike the GLM used for imputation for sensors with a small proportion of missing values, the model for neighbouring sensors requires transformation of the counts at the neighbours for imputation. As a consequence, it is more efficient to use `predict.glm()` inside the for loop as the neighbouring sensor counts have already been standardised when used for training the GLM. Again, the option `type = "response"` is specified to return predicted counts.

## Imputation Results

Firstly, we evaluate the initial model used to impute:
$$\text{where } \quad \mu_{\text{Sensor}, \text{HDay}, \text{Time}} \sim \text{Time}\times \text{HDay} \tag{1}$$

Note, this is also after the adjustments made to treat suspiciously long sequences of 0 values as `NA` in the actual counts. The simple GLM model is trained before this correction, while the improved model is trained on the adjusted data.

For the purposes of evaluating the imputation method at a location with a small proportion of missing values, we will first look at the counts from Southern Cross Station. At this sensor, the proportion of missing values (adjusted) is `r round(na_prop["Southern Cross Station"]*100,2)`Firstly, we evaluate the goodness of fit of the predictions made within the training period (01/01/2014 to 31/12/2016).

>>>>

---
knit: "bookdown::render_book"
---

# Abstract {-}

@@

<!-- 
The following line is required to re-set page numbering after preliminary material. Do not remove
-->
\clearpage\pagenumbering{arabic}\setcounter{page}{0}
